Automatically Proving Equivalence by Type-Safe Reflection

### 2017-03-27 paper-CICM-17/content/1-introduction.tex
added basic tactic for Nat and + assoc

When reading these definitions, one could wrongly think that the propositional equality is a simple wrapper of the judgemental equality. This is not the case : the propositional equality does not only contain the judgemental equality, because in these type theories, a principle of induction is associated with each inductive type. An inductive principle says that if a proposition holds for the base cases (ie, the constant constructors), and if it can be showed that when it holds for some terms then it will also hold for the bigger terms obtained by using the recursive constructors, then this proposition will hold for any term of this type. More formally, if a type $T$ is an inductive type with a constant constructor and a recursive constructor, ie, $T = 1 + T$ in type-theory, defined in idris as :

@@

The principles of induction can be used for proving any proposition, so they also work for the propositional equality. For example, we can prove that $n+0 = n$ for all $n$ by induction on the $Nat\ n$, even if $n+0 \not\equiv n$ with the usual definition of $+$, recursive on its first argument. That means that the axiom of induction has made the type $Id_{a}(x,y)$ bigger than it looks like : it contains terms that are not generated by $Refl$ (the only constructor of $Id_{a}(x,y)$), but that are added by the inductive principles. There are therefore things which are provably equal which unfortunately are not definitionally equal after unfolding all the definitions and after full reductions. The activity of proving equalities is therefore in these theories something which isn't automatically decidable by the type-checker in the very general case.

Proving that one term is equal to another is quite common
in formal certification. For example, when working with natural numbers, one might need to prove the lemma $\forall\ x, y:Nat,\ x*x + 3*x + x*y\ =\ x*(3+y+x)$. Or, when working with lists and concatenation of lists, one might need to prove the lemma $\forall\ l1,\ l2,\ l3:List A,\ (l1 ++ (l2 ++ l3)) ++ l4\ =\ (l1++l2) ++ (l3++l4)$. Currently, in the Idris programming language, these proofs must be completed by hand, and consist of a potentially long sequence of rewriting, where each step of rewriting uses one of the available properties. Proving this kind of auxiliary lemmas is time consuming and typically not reusable due to being very specific. These proofs are usually omitted ``on paper'' for everyday mathematics, but they are required by proof assistants, no matter how obvious they are to the human reader. These proofs can be an everyday routine when working in a dependently typed language, and especially when intensively indexing types over values in order to capture some logical properties, as we will show in the following subsection.

@@

We finally define the function $adc$ that adds two binary numbers and a carry flag. This addition works for two binary numbers expressed with the same number of bits, and it produces a result with one more bit. This result represents the value $c + x + y$, where $c$ represents the value of the input carry bit, and $x$ and $y$ represent the values of the two binary numbers given in input. Therefore, we would like to be able to write :

@@

The first pattern performs the addition between the values zero, zero and a carry. In this simple case, we simply have to produce the one-digit value $zero\ \#\ carry$. The second pattern is when the two binary numbers are respectively $(numx\ \#\ bX)$ and $(numy\ \#\ bY)$ where $numx$ and $numy$ are smaller binary numbers and $bX$ and $bY$ are bits added to the right of $numx$ and $numy$. In this case, we start by adding the least significant bits $bX$ and $bY$ with the $carry$, and this is performed by calling the function $addBit$. It will essentially produce 2 bits, the first one representing the carry of the result (called $carry0$), and the second one the least-significant bit of the result (called $lsb$). Once we have this intermediate result, we can call recursively $adc$ on the smaller terms $numx$ and $numy$ with $carry0$ as the carry that we need to propagate, and finally add $lsb$ to the right of this result.

Unfortunately, this straightforward definition is rejected by Idris' type-checker because the types mismatch for both of the two patterns. The result of the first line $adc\ zero\ zero\ carry$ is expected to have the type $Binary\ 1\ (plus\ (plus\ c\ 0)\ 0)$ but we provide a term of type $Binary\ 1\ (plus\ c\ (plus\ 0\ (plus\ 0\ 0)))$.
The problem is even worse for the second case where the expected index is \\
$(plus\ (plus\ c\ (plus\ bit2\ (plus\ v1\ (plus\ v1\ 0))))\ (plus\ bit\ (plus\ v\ (plus\ v\ 0))))$
while we're trying to provide a term indexed over \\
$plus\ vLsb\ (plus\ (plus\ (plus\ vCarry0\ v1)\ v)\ (plus\ (plus\ (plus\ vCarry0\ v1)\ v)\ 0))$.
We know that the definition of $adc$ we have given is correct, and it has \emph{provably} the expected type, but it does not have it \emph{immediately} or \emph{by reduction} : after full reductions\footnote{In dependently typed theories, the presence of types predicated on values creates the need of using reductions also inside of types}, the expected and provided types are still different, and idris can not figure out a proof of compatibility on its own. 
If we use provisional definitions[ref], the previous definition can be accepted by the system, but we now have two proof obligations $adc\_lemma\_1$ and $adc\_lemma\_2$ demanding proofs of equality between the two types, and they have to be provided by hand --here with a proof script-- :

@@

As we see, this kind of proof consists of a potentially long sequence of rewriting steps, where each step uses one of the available properties: neutral element, commutativity, associativity. Without some specific automation, this sequence of rewriting must be done by the programmer.
This is time consuming and more importantly, just a small change in the left or the right hand side of the equality may invalidate the proof. That means that a minor change in our datatypes, or in the definition of $addBit$ or $adc$ will require us to replace this proof by another one fulfilling the same task.
This kind of proof is therefore absolutely not reusable, because it performs rewritings for very specific terms.
These proofs become the everyday routine in any dependently-typed language, as soon as we index types over values in order to capture some logical properties.  As we've seen, when a type $T$ is indexed over $Nat$, the definition of a function which has codomain $T$ is likely to require a proof of equality between the produced index and the expected one.

@@

The principal novelty is in the approach that we follow, using a
new kind of \emph{type-safe reflection}.  Working by reflection for implementing tactics has been done several times, including for the implementation of a ring solver for the Coq proof assistant, but without
providing any guarantees, contrary to our type-safe reflection. We will
compare our approach with other implementations in Section~\ref{sect:AlternativeApproaches}.

>>>>

When reading these definitions, one could wrongly think that the propositional equality is a simple wrapper of the judgemental equality. This is not the case : the propositional equality does not only contain the judgemental equality, because in these type theories, a principle of induction is associated with each inductive type. If a type $T$ is an inductive type with a constant constructor and a recursive constructor, ie, $T = 1 + T$, defined in idris as :

@@

The principles of induction can be used for proving any proposition $P$, so they also work for the propositional equality. For example, we can prove that $n+0 = n$ for all $n$ by induction on the $Nat\ n$, even if $n+0 \not\equiv n$ with the usual definition of $+$, recursive on its first argument. That means that the axiom of induction has made the type $Id_{a}(x,y)$ bigger than it looks like : it contains terms that are not generated by $Refl$ (the only constructor of $Id_{a}(x,y)$), but that are added by the inductive principles. There are therefore things which are provably equal which unfortunately are not definitionally equal after unfolding all the definitions and after full reductions. The activity of proving equalities is therefore in these theories something which isn't automatically decidable by the type-checker in the very general case.

Proving that one term is equal to another is quite common
in formal certification. This is an everyday routine when working in a dependently typed language, and especially when intensively indexing types over values in order to capture some logical properties, as we will show in the following subsection.

@@

We finally define the function $adc$ that adds two binary numbers and a carry flag. This addition works for two binary numbers expressed with the same number of bits, and it produces a result with one more bit. This result represents the sum of the three concrete values of the inputs. Therefore, we want to write :

@@

Unfortunately, this straightforward definition is rejected by Idris' type-checker because the types mismatch for both of the two patterns. The result of the first line $adc\ zero\ zero\ carry$ is expected to have the type $Binary\ 1\ (plus\ (plus\ c\ 0)\ 0)$ but we provide a term of type $Binary\ 1\ (plus\ c\ (plus\ 0\ (plus\ 0\ 0)))$.
The problem is even worse for the second case where the expected index is \\
$(plus\ (plus\ c\ (plus\ bit2\ (plus\ v1\ (plus\ v1\ 0))))\ (plus\ bit\ (plus\ v\ (plus\ v\ 0))))$
while we're trying to provide a term indexed over \\
$plus\ vLsb\ (plus\ (plus\ (plus\ vCarry0\ v1)\ v)\ (plus\ (plus\ (plus\ vCarry0\ v1)\ v)\ 0))$.
We know that the definition of $adc$ we have given is correct, and it has \emph{provably} the expected type, but it does not have it \emph{immediately} or \emph{by reduction} : after full reductions\footnote{In dependently typed theories, the presence of types predicated on values creates the need of using reductions also inside of types}, the expected and provided types are still different, and Idris can not figure out a proof of compatibility on its own. 
If we use provisional definitions[ref], the previous definition can be accepted by the system, but we now have two proof obligations $adc\_lemma\_1$ and $adc\_lemma\_2$ demanding proofs of equality between the two types, and they have to be provided by hand --here with a proof script-- :

@@

As we see, this kind of proof consists of a potentially long sequence of rewriting steps, where each step uses one of the available properties: neutral element, commutativity, associativity. Without some specific automation, this sequence of rewriting must be done by the programmer.
This is time consuming and more importantly, just a small change in the left or the right hand side of the equality may invalidate the proof. That means that a minor change in our datatypes, or in the definition of $addBit$ or $adc$ will require us to replace this proof by another one fulfilling the same task.
These proofs become the everyday routine in any dependently-typed language, as soon as we index types over values in order to capture some logical properties.

@@

The principal novelty is in the approach that we follow, using a
new kind of \emph{type-safe reflection}.  Working by reflection for implementing tactics has been done several times, including for the implementation of a ring solver for the Coq proof assistant, but without the type-safe aspect. We will
compare our approach with other implementations in Section~\ref{sect:AlternativeApproaches}.
